Stability of variational quantum dynamics with Gaussian approximants

### 2014-04-22 paper.tex
big pre-backup push

\def\G{|Γ〉}
\def\A{|A〉}
\def\Gx{|Γ_χ〉}
\def\cite#1{$[\hbox{\tt #1}]$}
\def\T{^{\rm T}}

@@

The Gaussian wave packet is a ubiquitous idea in quantum mechanics.  Its discovery was inevitable: people were provoked by understanding that observables are essentially uncertain to study physical states in which they are normally distributed.  The Gaussian wavepacket also crops up as the quantum state which, placed in a quadratic potential, best approximates the harmonic oscillation of classical mechanics.  In this potential, there is a natural width for the packet, which ensures its shape remains constant during its motion.  This width can also be set by requiring equipartition of energy and the uncertainty principle.  Wavepackets with this natural width are coherent states.

It has also been a productive tool.  The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  If the problem is to find a wave function $|ψ(t)〉$ that evolves according to a Hamiltonian $H$, the quality of approximation can be measured by comparing the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$.  These are both vectors in Hilbert space, and can be compared by the norm of their difference.  The gaussian wave packets have often been used as variational approximants, because they are defined by their width and their central position and momentum, useful properties to determine for any quantum state.

The invention of the laser in 1963 called for a quantum theory of intense radiation.  Laser beams are hard to analyse because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.  The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.  Laser light is very coherent, so these expansions often represent it efficiently.  Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.  When very coherent matter, in the form of atomic Bose-Einstein condensates, became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics.  They have since been applied to other forms of condensed matter.  They have solved many problems, but are limited by intractable convergence issues.

@@

\beginsection{Continuous phase space expansions}

@@

@@

Coherent states are not orthogonal, but their brackets follow the law
$$〈β|α〉=e^{-½|β|²-½|α|²+β*α}=e^{-½|β-α|²+i\Im α*β},\tag\csip$$
whence
$$\bigl|〈β|α〉\bigr|²=e^{-|β-α|²}.$$
When kets are considered as vectors, the angle between coherent states satisfies
$$\cos θ=‖|α〉‖‖|β〉‖\cos θ=|〈β|α〉|=e^{-½|β-α|²},$$
and comparing Taylor series gives
$$θ=|β-α|+O(|β-α|²).$$
This has a simple geometric analogy where phase space is mapped locally onto the surface of the unit sphere, the coherent states are vectors from the origin to the surface, and $|α-β|$ is a spherical angle.  The analogy fails when the angles between vectors are no longer small; for example, it is misleading in regard to the relation between $|α〉$ and $∂_α|α〉$.

@@

Because we are after numerical methods, our goal is for expansions such as $\G f$ and $\A d$ to be numerically stable.  This means, for one thing, that a small change in $f$ should cause a small change in $\G f$.  The coherent states that compose $\G$ are overcomplete, so fixing the value of $\G f$ does not fix the value of $f$.  However, in order to compute with this expansion, we need to find a way of solving for $f$ such that a small change in $\G f$ causes a small change in the  $f$ that is the result of our method.

There is a precise way to state these conditions.  For the expansion $\G$ to be stable, its columns must form a frame\cite{1992-Daubechies-Ten}.  This means, in the continuous case, that there are constants $A$ and $B$ such that
$$ A‖|ψ〉‖²≤\int |〈γ|ψ〉|²\,dγ≤B‖|ψ〉‖²,$$
for all kets $|ψ〉$.  In the discrete case, the integral is replaced by a sum.  Because the operator $\G$ is linear, the ket $|ψ〉$ could be a difference $|φ₁〉-|φ₂〉$.  In this case, the right hand inequality guarantees that a small change in a ket will produce a small change in the components $〈γ|ψ〉$, which the left hand inequality says tht a small change in the components will produce a small change in the expanded ket.

(Talk about dual frames, minimal norm expansions.)

Many of the above results for the operator $\G$ are true of expansions over frames in general.  We will treat the finite case, because every frame will eventually have to be represented on a computer.  So suppose the set of kets $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a frame for some subspace of Hilbert space.

The theory of frames is traditionally presented in terms of the frame operator,
$$S=∑_{r=1}^R |a_r〉〈a_r|={\rm Tr}(|A〉〈A|).$$
In the case that the columns of $|A〉$ are a frame, $S$ is invertible.  The set of vectors
$$|F〉=S^{-1}|A〉=\pmatrix{S^{-1}|a₁〉&⋯&S^{-1}|a_R〉}
	=\pmatrix{|φ₁〉&⋯&|φ_R〉}
$$
is also a frame, and is known as the dual frame of $|A〉$.  It has the property that $|A〉〈F|$ is the identity operator on the space.

@@

%\topinsert \XeTeXpicfile gweight.pdf width \hsize 
The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.  The points $x_i$ are normally distributed (top), or evenly spaced (bottom).  Note how nearby points tend to cancel each others' samples.  I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.

The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.  This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.  Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.  Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.  When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.  This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.  So one solution  would be to find the best expansion of $e^{-x²}$, instead of 1, then adjust the weight for $x_i$ by either $e^{-x_i²}$ or $〈e^{-(x-x_i)²},e^{-x²}〉$.
%\endinsert

@@

\beginsection{Regularisation of the discrete expansion problem}
%% Picard plots

@@

\midinsert %\XeTeXpicfile tysv.pdf width \hsize
Figure 3 \endinsert

@@

%\topinsert \XeTeXpicfile geometry-1.pdf
A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.  As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.  This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
%\endinsert

%\topinsert \XeTeXpicfile cancel.pdf width \hsize
The vacuum state was expanded over two coherent states, with amplitudes $±h$.  The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.  The norm of the expansion vector, and the residual of the expansion, are graphed above as black lines.  As the amplitudes converge, the residual decreases—the leftmost phase space plot is very nearly exact—while the norm of the expansion stays around 1.  This coincides with the amplitudes passing through bright areas of the images, where the entire expansion has support.

>>>>

\def\G{|Γ〉}
\def\A{|A〉}
\def\Gx{|Γ_χ〉}
\def\cite#1{$[\hbox{\tt #1}]$}
\def\define#1{#1}
\def\T{^{\rm T}}

@@

The Gaussian wave packet is a ubiquitous idea in quantum mechanics.  Its discovery was inevitable: people were provoked by understanding that observables are essentially uncertain to study physical states in which they are normally distributed.  The Gaussian wavepacket also crops up as the quantum state which, placed in a quadratic potential, best approximates the harmonic oscillations of classical mechanics.  The potential sets a natural width for the packet, which causes its shape to stay constant as it moves.  This width is also characterised by equipartition of energy and the uncertainty principle.  Wavepackets with this natural width are called coherent states.

Wave packets have been a productive tool.  The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  If the problem is to find a wave function $|ψ(t)〉$ that evolves according to a Hamiltonian $H$, the quality of approximation can be measured by comparing the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$.  These are both vectors in Hilbert space, and can be compared by the norm of their difference.  Gaussian wave packets have often been used as variational approximants, because they are defined by their width and their central position and momentum, useful properties to determine for any quantum state.

The invention of the laser in 1963 called for a quantum theory of intense radiation.  The difficulty of analysing a laser beam stems from the size of its quantum description, up to $10^{18}$ photons distributed across thousands of modes.  This was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.  Laser light is very coherent, so these expansions often represent it efficiently, converging with a reasonable number of states in the superposition.  Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.  When very coherent matter, in the form of atomic Bose-Einstein condensates, became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics.  They have since been applied to other forms of condensed matter.  They have solved many problems, but are limited by intractable convergence issues.

@@

One modification would be to use a norm derived from lower order moments of the creation and annihilation operators\cite{Drummond-private}.  In this case, there would not be a direct link between the variational parameters and a state in Hilbert space.  Instead, the superposition of coherent states would be a way to record moments of the system, which would be propagated so that the rates of change of the moments had the expected values.  It is unclear why recording the moments indirectly, by an ensemble of coherent amplitudes, would be better than recording the moments themselves.  Several methods already exist for propagating the moments of a quantum optical system
: these usually assume that some higher-order moments factor in terms of lower-order ones, in order to obtain a system of ordinary differential equations for a finite set of moments.  This assumption is at least explicit, and some equivalent assumption would have to be made implicitly if the moments were derived from an ensemble of coherent amplitudes.  It is unclear why this would be a better approximation to quantum mechanics than the differential equations obtained by assuming the moments factor.

\beginsection{Coherent states}

TODO: Define “phase space” at some point.

@@

Coherent states are not orthogonal, but their brackets follow the law
$$〈β|α〉=e^{-½|β|²-½|α|²+β*α}=e^{-½|β-α|²+i\Im α*β},\tag\csip$$
whence
$$\bigl|〈β|α〉\bigr|²=e^{-|β-α|²}.\tag\csolap$$
When kets are considered as vectors, the angle between coherent states satisfies
$$\cos θ=‖|α〉‖‖|β〉‖\cos θ=|〈β|α〉|=e^{-½|β-α|²},$$
and comparing Taylor series gives
$$θ=|β-α|+O(|β-α|²).$$
This has a geometric analogy.  The normalised kets form a hypersphere in ket space.  A two-dimensional manifold of normalised kets, such as the coherent states, must be locally similar to the unit sphere.  It turns out that phase space is mapped into ket space without local distortion: the relation between a coherent ket and its neighbours is as if phase space were the surface of a sphere.  This is only the case when the angles between vectors are small; for example, it is misleading in regard to the relation between $|α〉$ and $∂_α|α〉$.

Equation~\csip\ has a consequence for expansions.  Suppose a coherent state $|β〉$ is expanded in the form in Equation~\Ad, with components such that $|β-α_r|>δ$, and error $‖|β〉-|ψ〉‖<ε$.  Then $|〈β|ψ〉|>1-ε$, but $|〈β|ψ〉|=|∑〈β|α_r〉d_r|≤e^{-{δ²\over 2}}∑|d_r|$.  Therefore, as the expanded state moves away from the components, the coefficients must grow as $e^{δ²\over 2}$.  The error condition requires that $‖|ψ〉‖<1+ε$, so these components must almost entirely cancel each other out.  This bodes badly for the numerical stability of such expansions: the details are explored below.

TODO: say that the default norm in ${\bf C}^n$ is the 2-norm.

In order to support numerical methods, the expansion $|ψ〉=\A d$ needs to be numerically stable.  This means, most obviously, that changing $d$ by a small amount should cause a small change in $\A d$.  This is clearly the case.  The coherent states that compose $\A$ are normalised, so clearly $‖\A d‖≤R‖d‖$, whence $‖\A‖≤R$.

TODO: tighten this bound (can we get $‖\A‖≤\sqrt{R}$?)

The opposite condition, that a small change in $|ψ〉$ should cause a small change in $d$, is trickier.  Finite sets of coherent states are linearly independent in the mathematical sense, so the equation $\A d₀=0$ has no solution, and the operator $\A$ is invertible.  This is not true in the numerical sense: given a typical set of amplitudes $\{α_r\}$, there will be vectors $d₀$ such that $\A d₀$ lies within rounding error of zero.  We can add any multiple of $d₀$ to $d$ without changing the computed $|ψ〉$ at all, so the problem of finding $d$ is utterly unstable.  The best we can hope for is to find a specific expansion that is numerically stable.

As a first step, we can require that the components $〈A|ψ〉$ are stable.  The conventional way to formulate that is to require there are finite constants $P$ and $Q$ such that
$$ P‖|ψ〉‖²≤‖〈A|ψ〉‖²≤Q‖|ψ〉‖²,$$
for all kets $|ψ〉$.  In this case, the set comprising kets $|α_r〉$ is known as a \define{frame}.  Strictly, the mapping $〈A|$ from kets to components is called the frame operator; this paper will often refer to $\A$ as a frame, meaning that its columns form a frame.  The constant $P$ is called the lower frame bound, and $Q$ the upper frame bound.  It will turn out that, in order to form stable expansions, these bounds ought to be close.  The ideal case, where they are equal, is known as a tight frame.

There is a complication that we'll ignore for now.  The state $|ψ〉$ lies in an infinite dimensional Hilbert space, in which there are many directions that are orthogonal to every $|α_r〉$.  When $|ψ〉$ points along such a direction, $〈A|ψ〉=0$.  So no finite set of kets is a frame for Hilbert space, but only for some subspace.  We will discuss this in detail below.

It can be shown that if $\A$ is a frame, the operator $\A〈A|$ is invertible\cite{1992-Daubechies-Ten}.  In this case, the set of kets
$$|F〉=(\A〈A|)^{-1}|A〉=\pmatrix{(\A〈A|)^{-1}|a₁〉&⋯&(\A〈A|)^{-1}|a_R〉}
	=\pmatrix{|φ₁〉&⋯&|φ_R〉}
$$
is also a frame, called the dual frame of $|A〉$.  These kets have the property that $|A〉〈F|=|F〉〈A|=1$, the identity operator on the space spanned by the frame.  In other words, the components of $|ψ〉$ along the directions of the dual frame $|F〉$ are the coefficients of an expansion over the frame $|A〉$, and visa-versa.  If $|α_r〉$ formed an orthonormal basis, we would have $|F〉=|A〉$, and $|A〉〈A|=1$.  In this sense, frames are a generalisation of orthonormal bases.  For a tight frame, $|F〉={1\over P}|A〉$, and the analogy is especially close.

\beginsection{Phase space as a Gabor frame}

@@

@@

@@

\topinsert \XeTeXpicfile gweight.pdf width \hsize 
The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.  The points $x_i$ are normally distributed (top), or evenly spaced (bottom).  Note how nearby points tend to cancel each others' samples.  I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.

The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.  This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.  Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.  Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.  When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.  This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.  So one solution  would be to find the best expansion of $e^{-x²}$, instead of 1, then adjust the weight for $x_i$ by either $e^{-x_i²}$ or $〈e^{-(x-x_i)²},e^{-x²}〉$.
\endinsert

@@

\beginsection{The discrete variational frame}

In the variational problem, the derivative $-iH|ψ(z)〉$ is expanded over the frame $|Dψ(z)〉$.  For the coherent state approximant of Equation~?,
$$ |Dψ(z)〉=\pmatrix{1 & a†}|ψ(z)〉.$$
It follows that
$$ 〈∂₁ψ|∂₁ψ〉=〈ψ|ψ〉=e^{2\Re φ+|α|}
	\qquad 〈∂₂ψ|∂₂ψ〉=(1+|α|²)〈ψ|ψ〉,$$
and
$$\qquad 〈∂₁ψ|∂₂ψ〉=e^{2\Re φ+|α|}〈α|a†|α〉=α*〈ψ|ψ〉.$$
The angle $θ$ between $|∂₁ψ〉$ and $|∂₂ψ〉$, such that
$$ |〈∂₁ψ|∂₂ψ〉|=‖|∂₁ψ〉‖‖|∂₂ψ〉‖\cos θ,$$
satisfies $\cot θ=|α|$.  The partial derivatives are orthogonal for the vacuum state $α=0$, but become nearly parallel as $|α|$ increases.  This will increase the condition number of $|Dψ〉$.

\beginsection{Regularisation of the discrete problems}
%% Picard plots

@@

\midinsert \XeTeXpicfile tysv.pdf width \hsize
Figure 3 \endinsert

@@

\topinsert \XeTeXpicfile geometry-1.pdf
A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.  As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.  This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
\endinsert

\topinsert \XeTeXpicfile cancel.pdf width \hsize
The vacuum state was expanded over two coherent states, with amplitudes $±h$.  The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.  The norm of the expansion vector, and the residual of the expansion, are graphed above as black lines.  As the amplitudes converge, the residual decreases—the leftmost phase space plot is very nearly exact—while the norm of the expansion stays around 1.  This coincides with the amplitudes passing through bright areas of the images, where the entire expansion has support.
